---
title: "How to create your own movie recommender system in R using collaborative filtering technique"
author: "Adeyemi Arogunmati"
date: "June 13, 2016"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

```{r load properties, echo=FALSE}
load(file="data_dimensions.RData")
load(file="data_averages.RData")
load(file="total_ratings.RData")
tratings <- as.character(totalRatings)
nusers   <- as.character(dimRatings[2])
nmovies  <- as.character(dimRatings[1])
average  <- as.character(round(totalRatings/dimRatings[2]))
```

## Introduction
Ever wonder how automated movie recommender systems work: systems where you rate a few movies, then other movies are recommended to you based on your inferred preferences? Well, I've put together a documentation on how I built one of my own in R without the use of black-box algorithms or tools. I used the collaborative filtering technique. You can modify/customize my workflow to suit your needs.

## Getting Data
To begin, you will need access to a database of movies, all of which have been rated by other people. For simplicity, let's call these people users. The users only have to rate some movies not all movies, but all movies have to be rated by some users. Fortunately for us, these sorts of databases exist. I will use the publicly available [MovieLens data set collected by GroupLens Research](!http://grouplens.org/datasets/movielens/).
Because I want to use a data set containing movies as recently released as posible, I will use the [MovieLens Latest Dataset, ml-latest-small.zip (download date: Jun 13, 2016)](!http://grouplens.org/datasets/movielens/latest/). The ml-latest-small data set contains `r I(tratings)` ratings across `r I(nmovies)` movies provided by `r I(nusers)` users and was generated on January 11, 2016. The average number of ratings per user is `r I(average)`.

For files are included in the data set: `links.csv`, `movies.csv`, `ratings.csv` and `tags.csv`. I am most interested in the `movies.csv` and `ratings.csv` files. 

User ratings are contained in the `ratings.csv` file which is a csv file with columns representing 

- userId: user identification tag 
- movieId: movie identification tag
- rating: user's rating (1-5 stars) of movie with specified movieId
- timestamp: timestamp of when movie was rated

Movie titles are contained in the `movies.csv` file which is a csv file with columns representing

- movieId: movie identification tag
- title: movie title
- genres: movie genre

Import data to project:
```{r import data to project, eval=FALSE}
movies  <- read.csv("data/ml-latest-small/movies.csv",  header=T)
ratings <- read.csv("data/ml-latest-small/ratings.csv", header=T)
```


## Data Wrangling
After obtaining the movie ratings data set, the next step is to transform it into the right format. For our purpose, the right format is a matrix with rows representing individual movies and columns representing individual users. The cells of this matrix contains the rating of a movie represented by that row by the user represented by that column. This transformation can be done in R using nested for loops. Note that if done in R, the process will take a long time to complete. I wrote a c++ code to accomplish the task in seconds. This c++ code, `main.cpp` is located in the `data_and_codes/` directory. 

We also need to create a matrix that indicates which cells have ratings in them. We will use the value 1 to represent cells with ratings and 0 otherwise.

R code:
```{r data to matrix, eval=FALSE}
# get maximum user id number
userId_max  <- max(ratings$userId)
# get maximum movie id number
movieId_max <- max(ratings$movieId)

# create matrix with number of rows equal to maximum number
# of movies and columns equal to maximum number of users 
ratingsMatrix <- matrix(0,length(1:movieId_max),length(1:userId_max))
ratedMatrix   <- matrix(0,length(1:movieId_max),length(1:userId_max))

# loop through data and fill matrices
for (i in 1:movieId_max) {
    for (j in 1:userId_max) {
        rown <- ratings[ratings$movieId == i & ratings$userId == j,]
        if (nrow(rown) > 0) {
            ratingsMatrix[i,j] <- rown$rating
            ratedMatrix[i,j]   <- 1
        }
    }
}

# get ids of rated movies
movieIdColumn <- seq(1:nrow(ratingsMatrix))
```

If c++ code is used, read output files:
```{r read c plus plus output files, eval=FALSE}
ratingsMatrix <- read.csv("ml-latest-small_ratingsMatrix.csv", header=F)
ratedMatrix   <- read.csv("ml-latest-small_ratedMatrix.csv",   header=F)

# convert data frames to matrices
ratingsMatrix <- as.matrix(ratingsMatrix)
ratedMatrix   <- as.matrix(ratedMatrix)
movieIdColumn <- seq(1:nrow(ratingsMatrix))
```

Because we created a matrix large enough to have row numbers as large as the largest movieId, there are lots of empty rows or rows with only zeros in our matrices. We should remove these rows.

```{r remove empty rows, eval=FALSE}
# remove empty rows
ratingsMatrixFinal <- ratingsMatrix[as.logical(rowSums(ratingsMatrix != 0)), ]
ratedMatrixFinal   <- ratedMatrix  [as.logical(rowSums(ratingsMatrix != 0)), ]
movieIdColumnFinal <- movieIdColumn[as.logical(rowSums(ratingsMatrix != 0))  ]
```


## Add My Ratings
After creating a matrix of ratings, I added my ratings of 53 movies. I identified movies that I would rate 1, 2, 3, 4, and 5 stars. You can find my ratings in the file `myratings_mod.csv` in the directory `data_and_codes/`. This step can be done prior to data wrangling, but I chose to do it after. Results will be the same.
```{r adding my ratings, eval=FALSE}
# import csv file containing my ratings
myratingsraw <- read.csv("data_and_code/myratings_mod.csv",header=T)

# find which rows represent the movie I rated
findRow <- function(item,vec) which(vec==item)
myratingsrows <- unlist(
    sapply(myratingsraw$movieId,
           findRow,movieIdColumnFinal))

myratings <- numeric(nrow(Y))

for (i in 1:length(myratingsrows)) {
    r <- myratingsrows[i]
    myratings[r] <- myratingsraw$rating[i]
}

# set ratingsMatrixFinal to Y and ratedMatrixFinal to R
Y <- ratingsMatrixFinal
R <- ratedMatrixFinal

# append my ratings to Y and R
myY <- cbind(myratings, Y)
myR <- cbind(as.integer(myratings != 0), R)
    

# compute the mean of ratings for each movie
# we will subtract this value from all ratings 
# before running collaborative filtering
myY_sum <- rowSums(myY)
myR_sum <- rowSums(R)
Ymean   <- y_sum/r_sum
```

## Collaborative Filtering Cost Function and Gradient
Since we will be using gradient descent in solving our collaborative filtering algorithm, we need to define functions that compute our cost function and gradient. This function takes as input, a regularization parameter lambda, a matrix of movie rating indicators R, a matrix of movie coefficients Theta, and a matrix of user coefficients X. We will use collaborative filtering to compute the matrices X and Theta. Since we don't know these at the onset, we will initialize them with random numbers.

```{r collaborative filtering algorithm, eval=FALSE}
collaborativeFilter <- function (X, Y, Theta, R, lambda) {
    # 1. cost function
    J <- 1/2 * ( (X%*%t(Theta)) - Y )^2 * R;
    
    reguTheta <- lambda/2 * sum(Theta^2) # Theta regularization
    reguX     <- lambda/2 * sum(X^2)     # X regularization
    
    J <-  sum(J) + reguTheta + reguX
    
    # 2. gradients
    X_grad     <- ((X%*%t(Theta)-Y)  * R) %*% Theta + lambda * X;
    Theta_grad <- t((X%*%t(Theta)-Y) * R) %*% X     + lambda * Theta;
    
    list(J,X_grad,Theta_grad)
}
```

## Execution
We now have everything we need to run our algorithm. Let's put them together.

### 1. Initialize matrices, vectors, and scalars
```{r initializa values, eval=FALSE}
# set Y to myY and R to myR
Y <- myY
R <- myR

# mean normalize Y. i.e. subtract mean ratings
for (i in 1:ncol(Y)) Y[,i] <- Y[,i] - Ymean

# set scalars
nmovies   <- nrow(Y) # number of movies
nusers    <- ncol(Y) # number of users
nfeatures <- 20

lambda <- 1  # regularization parameter set to 1
niter  <- 30000 # number of iterations
alpha  <- 0.00004 # gradient descent step parameter

costvec <- numeric(niter) # initialize vector of cost computation

set.seed(1234) # set random number generator seed to 1234

# initialize X to matrix of uniformly distributed random numbers
X     <- matrix(runif(nmovies*nfeatures),nmovies,nfeatures)

# initialize Theta to matrix of uniformly distributed random numbers
Theta <- matrix(runif(nusers*nfeatures),nusers,nfeatures)
```


### 2. Run collaborative filtering with gradient descent
```{r gradient descent, eval=FALSE}
for (i in 1:niter) {
    collabfilt <- collaborativeFilter(X,Y,Theta,R,lambda)
    
    costvec[i] <- collabfilt[[1]]
    X_grad     <- collabfilt[[2]]
    Theta_grad <- collabfilt[[3]]
    
    X     <- X     - alpha * X_grad
    Theta <- Theta - alpha * Theta_grad
}
```


### 3. Extract movie recommendations
```{r load RData file, cache=TRUE, echo=FALSE}
load(file="my_ratings_normalize_cofi_06-13-2016.RData")
```

```{r extract recommendations}
# get movie titles in vector
findMovieTitles <- function(item,vec) which(vec==item)
mymovierows <- unlist(
    sapply(movieIdColumnFinal,findMovieTitles,movies$movieId))

# compute my predictions and add column with movie titles
p <- X %*% t(Theta) + Ymean

# create data frame with predictions and movie titles
mypred <- data.frame(movieTitle=movies$title[mymovierows],
                     movieId=movieIdColumnFinal,
                     rating=round(p[,1],1)
                     )

# sort prediction by rating
mypred_sort <- mypred[with(mypred, order(-rating)), ]

# add movie release year to prediction data frame
getMovieYear <- function(text) {
    y1  <- gsub("[^0-9]","",gsub("[^0-9\\(\\)]","",text))
    len <- nchar(y1)
    y2  <- substr(y1, len-3, len)
}
mypred_sort$movieYear <- unlist(sapply(mypred_sort$movieTitle,getMovieYear))
mypred_sort$movieYear <- as.integer(mypred_sort$movieYear)

# sort prediction by release year, then rating
mypred_sort <- mypred_sort[with(mypred_sort, order(-movieYear,-rating)), ]
```


### 4. Display recommendations for top 10 movies released in 2013, 2014 and 2015
Show recommendations of movies that I have not rated but predicted to be rated highly by me.
```{r recommend 1, echo=FALSE, eval=FALSE}
data2015 <- mypred_sort[mypred_sort$movieYear==2015,]
data2014 <- mypred_sort[mypred_sort$movieYear==2014,]
data2013 <- mypred_sort[mypred_sort$movieYear==2013,]
recommend <- rbind(data2015[1:10,],data2014[1:10,],data2013[1:10,])
names(recommend) <- c("Movie Title","Movie Id","Predicted Rating","Release Year")
rownames(recommend) <- c()
kable(recommend[,c(1,3)])
```

```{r recommend 2, echo=FALSE, cache=TRUE}
# get rows of movies I rated
r <- numeric(0)
aa <- as.character(myratingsraw$movieId)
for (i in 1:nrow(mypred_sort)) {
    t <- mypred_sort$movieId[i]
    b <- as.character(t)
    b <- paste0(c("^",b,"$"),collapse="")
    for (j in 1:length(aa)) {
        a <- aa[j]
        if (grepl(b,a)) {
            r <- rbind(r,i)
        }
    }
}
r <- as.integer(r)

# remove those rows from the predicted ratings
mypred_sort2 <- mypred_sort[-r,]

# select top 10 from 2013, 2014, and 2015
data2015 <- mypred_sort2[mypred_sort2$movieYear==2015,]
data2014 <- mypred_sort2[mypred_sort2$movieYear==2014,]
data2013 <- mypred_sort2[mypred_sort2$movieYear==2013,]
recommend <- rbind(data2015[1:10,],data2014[1:10,],data2013[1:10,])
names(recommend) <- c("Movie Title","Movie Id","Predicted Rating","Release Year")
rownames(recommend) <- c()

# display recommendations
kable(recommend[,c(1,3)])
```

These are definitely movies I would enjoy :)

## Conclusions
To conclude, we have built a movie recommender system based on collaborative filtering technique. This movie recommender system uses provided movie ratings to make predictions on movie ratings of new users. We added my ratings of 53 movies and used the recommender system to make recommendations based on my computed movie preferences. A visual assessment of the generated list shows to my that the recommendation system we built works!



